A Large-scale Study of Polymorphism in R

### 2019-03-30 paper/main.tex
updating prose in results, getting through L0

\AT{This is the (partial) data for the new layout.
We need to decide how to tackle L3.}

Figure~\ref{fig:L1argcounts} has arg counts for L1.

Figure~\ref{fig:L1funcounts} has fun counts for L1.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 29 18:10:23 2019
\begin{table}[ht]
\label{fig:L1argcounts}
\centering
\begin{tabular}{lrr}
  \hline
Type & Count & Percentage \\ 
  \hline
Full Monomorphic & 784K & 85.76 \\ 
  Monomorphic in Type & 845K & 92.46 \\ 
  Monomorphic in Class & 827K & 90.41 \\ 
  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
  Total Seen & 914K & --- \\ 
   \hline
\end{tabular}
\caption{typeof type system (L1) (arguments)}
\end{table}
% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 29 18:10:23 2019
\begin{table}[ht]
\label{fig:L1funcounts}
\centering
\begin{tabular}{lrr}
  \hline
Type & Count & Percentage \\ 
  \hline
Full Monomorphic & 144K & 66.54 \\ 
  Monomorphic in Type & 173K & 80.16 \\ 
  Monomorphic in Class & 163K & 75.21 \\ 
  Monomorphic in Attribute Pattern & 161K & 74.37 \\ 
  Total Seen & 216K & --- \\ 
   \hline
\end{tabular}
\caption{typeof type system (L1) (functions)}
\end{table}

Table~\ref{tab:L1toppoly} has the top polymorphic signatures for L1.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 29 18:13:17 2019
\begin{table}[ht]
\label{tab:L1toppoly}
\centering
\begin{tabular}{lrr}
  \hline
type & count & perc \\ 
  \hline
double, integer & 19799 & 28.73 \\ 
  double, list & 5324 & 7.72 \\ 
  double, NULL & 5152 & 7.48 \\ 
  character, NULL & 3682 & 5.34 \\ 
  double, logical & 3388 & 4.92 \\ 
  character, double & 2842 & 4.12 \\ 
  list, NULL & 2799 & 4.06 \\ 
  character, list & 1836 & 2.66 \\ 
  character, integer & 1663 & 2.41 \\ 
  character, double, integer & 1362 & 1.98 \\ 
   \hline
\end{tabular}
\caption{Top polymorphic signatures for the L1 type system.}
\end{table}

@@

Having built an understanding of our data collection and analysis
pipeline, we will turn our attention to the results of our analysis.  We
approach the data with the purpose of informing the design of a static type
system for R.

The logical ``baseline'' for the design of a static type system for R is
promotion of the language's dynamic types (in the sense of \code{typeof}) to
static types, but there are some clear limitations to this approach.  For
example, \code{typeof} makes no distinction between vectors and scalars, as
indeed there is no such distinction in R.  But that doesn't mean that there
{\it couldn't} be a distinction, as an all-new runtime environment design
for R might benefit from e.g. not needing to vectorize scalar values.

>>>>

Having built an understanding of our data collection and analysis pipeline, we will turn our attention to the results of our analysis.

The logical ``baseline'' for the design of a static type system for R is promotion of the language's dynamic types (in the sense of \code{typeof}) to static types; we call this type system L0.
Such a type system exactly reflects the types of values in the R runtime environment, as the \code{typeof} an object is in direct correspondence with the runtime type tag of the value.

In this section, we will begin by breaking down our corpus of data with L0 in mind.
From this perspective, we will see how current R programmers see types in the language, and we will leverage our newfound insight to design a type system which should be relevant to them.

Without further ado, let's look at the data, first focusing on a broad overview of the polymorphism in L0.
Figure~\ref{fig:L1argcounts} shows the polymorphicity of argument positions in R function.
From there, we build up a picture of the polymorphicity of functions as a whole, and the data for this function polymorphism can be found in Figure~\ref{fig:L1funcounts}.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 29 18:10:23 2019
\begin{table}[ht]
\label{fig:L1argcounts}
\centering
\begin{tabular}{lrr}
  \hline
Type & Count & Percentage \\ 
  \hline
Full Monomorphic & 784K & 85.76 \\ 
  Monomorphic in Type & 845K & 92.46 \\ 
  Monomorphic in Class & 827K & 90.41 \\ 
  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
  Total Seen & 914K & --- \\ 
   \hline
\end{tabular}
\caption{typeof type system (L0) (arguments)}
\end{table}
% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 29 18:10:23 2019
\begin{table}[ht]
\label{fig:L1funcounts}
\centering
\begin{tabular}{lrr}
  \hline
Type & Count & Percentage \\ 
  \hline
Full Monomorphic & 144K & 66.54 \\ 
  Monomorphic in Type & 173K & 80.16 \\ 
  Monomorphic in Class & 163K & 75.21 \\ 
  Monomorphic in Attribute Pattern & 161K & 74.37 \\ 
  Total Seen & 216K & --- \\ 
   \hline
\end{tabular}
\caption{typeof type system (L0) (functions)}
\end{table}

From Figure~\ref{fig:L1argcounts}, we glean that argument positions aren't very polymorphic in L0.
An argument which is "Monomorphic in Type" is monomorphic w.r.t. \code{typeof}, in that it was only ever observed to have one \code{typeof}.
Similarly, an argument is "Monomorphic in Class" if it is only ever called with one {\tt class}.
As we said earlier, attributes are a little more complicated, and an argument which is "Monomorphic in Attribute Pattern" was only ever observed to have one attribute pattern (and that pattern may itself contain many attributes).

As for the polymorphism of functions as a whole, we glean from Figure~\ref{fig:L1funcounts} that a sizeable number of functions are polymorphic.
Functions are said to be polymorphic if at least one of their arguments or return are polymorphic.
\AT{Say more?}

At present, we will only concern ourselves with \code{typeof} polymorphism, and we will later worry about classes and attributes.
To get an idea of what kinds of polymorphism occur, we will look at the most commonly occurring polymorphic argument signatures.
Table~\ref{tab:L1toppoly} has the top polymorphic signatures for L0.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 29 18:13:17 2019
\begin{table}[ht]
\label{tab:L1toppoly}
\centering
\begin{tabular}{lrr}
  \hline
type & count & perc \\ 
  \hline
  double, integer & 19,799 & 28.73 \\ 
  double, list & 5,324 & 7.72 \\ 
  double, NULL & 5,152 & 7.48 \\ 
  character, NULL & 3,682 & 5.34 \\ 
  double, logical & 3,388 & 4.92 \\ 
  character, double & 2,842 & 4.12 \\ 
  list, NULL & 2,799 & 4.06 \\ 
  character, list & 1,836 & 2.66 \\ 
  character, integer & 1,663 & 2.41 \\ 
  character, double, integer & 1,362 & 1.98 \\ 
   \hline
\end{tabular}
\caption{Top polymorphic signatures for the L0 type system.}
\end{table}

Some entries in the Table~\ref{tab:L1toppoly} are clear (e.g., double, integer polymorphism is unsurprising), others are strange (e.g., character, double), but in fact they all have some explanation.

	\item \{double, integer\}: in R, doubles and integers are effectively interchangeable, as casting from one to the other is straightforward and automatically performed when appropriate.
	For example, \code{c(1, 2)[1.9] == 1}.
	We can easily consolidate these into one type, say a real number type.

	\item \{double, list\} and \{character, list\}: at first glance, these type signatures are suspect.
	Why would so many programmers want to pass a double or a list to some function?
	The key insight here is that doubles and characters are {\it vectorized} in R, meaning that a double is really a vector of doubles.
	Thus, these seemingly strange signatures are revealed to be entirely reasonable: these signatures are likely to describe arguments which are either vectors or lists.

	\item \{X, NULL\}: in most languages, NULL is an inhabitant of all types, and we see that this is not the case in R.
	That said, there is a pattern of many functions having default or optional arguments, and in many of these cases NULL is the default value.
	This is a distinction that might be worth collapsing, and further this indicates that some annotation for optional arguments would be useful.

	\item \{double, logical\}: the first values of type logical that come to mind are TRUE and FALSE, which does little to explain this strange signature.
	Indeed, the key here is an implementation detail of R's: recall that NAs inhabit every type.
	As it happens, the default type of NA is logical (e.g., \code{typeof(NA) == "logical"}), and this is likely what is being seen here.
	We will control for this by distinguishing NAs with their own type, {\tt raw\_NA}, but we will get to that in time.

	\item \{character, double\}, \{character, integer\} and \{character, double, integer\}: yet more strange signatures.
	These signatures are a little more subtle to explain.
	For one, indexing lists and vectors can be done with either strings (i.e., to access a named index) or with numbers.
	Another option is dates: R has many ways to present dates, and strings and numbers feature among them.
	Finally, many \AT{get numbers} functions in R return a string if an error occurs during execution.

@@

\AT{Limitations of L0.}
... but there are some clear limitations to this approach.  For
example, \code{typeof} makes no distinction between vectors and scalars, as
indeed there is no such distinction in R.  But that doesn't mean that there
{\it couldn't} be a distinction, as an all-new runtime environment design
for R might benefit from e.g. not needing to vectorize scalar values.
